NMRlipids Databank makes data-driven analysis of biomembrane properties accessible for all

### 2023-03-20 main.tex

Cellular membranes contain hundreds of different types of lipid molecules that regulate the membrane properties, morphology, and biological functions~\cite{vanmeer08,Lorent:2020a,Slatter:2016a}. Membrane lipid composition is implicated in diseases, such as cancer and neurodegenerative disorders, and therapeutics that affect membrane compositions are emerging~\cite{torres21}. However, biomembranes are often difficult to study experimentally because at biological conditions they are in disordered fluid state, and sample preparation can be arduous due to the complicated phase behaviour and rich interactions between lipids and other biomolecules. For those reasons, the detailed connections between complex lipid interactions and biological functions taking place in or around membranes remain poorly understood. Molecular dynamics (MD) simulations have been particularly useful in understanding membrane systems, although their accuracy has often been compromised by artefacts such as the quality of model parameters~\cite{antila22b,gupta22}. Presently, the accuracy of models is becoming increasingly important as researches are progressing from simulations of individual molecules to simulating whole organelles or even cells using interdisciplinary approaches~\cite{johnson15,thornburg22,gupta22}. Such systems exhibit complex emergent behaviour making inaccuracies more difficult to detect and accumulation of even modest errors may have major impact on the conclusions.

@@

\subsection{NMRlipids databank-API: programmatic access to the MD simulation data}
The NMRlipids databank-API, available at \url{www.github.com/NMRLipids/Databank}, provides programmatic access to all simulation data in the NMRlipids databank. This enables wide range of novel data-driven applications from construction of machine learning models that predict membrane properties to automatic analyses of virtually any property from all simulations in the databank. 
%Wide range of novel analyses taking the advantage of programmatic access to unprecendented amount MD simulation data can be performed with the tools delivered by the NMRlipids databank-API. 
Practical implementation of such analyses is illustrated in the flowchart in Fig.~\ref{fig:quality}D. First, raw data of each simulation can be accessed by iterating through the README.yaml files in the {\it databank layer} that contain links pointing to the locations of simulation trajectories and other relevant files. Simulations can be then automatically analyzed with the help of README.yaml and mapping files that affiliate specific naming conventions in each simulation to the universal molecule and atom names. Finally, the analysis results are stored using the same structure as in the {\it databank layer}.

@@

\subsection{Applications of the NMRlipids databank}
\subsubsection{Selection of simulation parameters using the NMRlipids databank: Best models for most abundant neutral membrane lipids}
%Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19,??}, but the lack of universal quality measure for membrane simulations complicates the estimations of reliability of simulations. 
%
%Understanding of how closely simulations resemble the reality is crucial for selecting best models for specific applications, assessing the reliability and significance of MD simulation results, and using atomistic simulations in parametrization of coarse grained models. However, undefined universal quality measures for MD simulations and the lack of a prevailing force field that would perform best in all simulations has led to many controversial results in membrane simulations~\cite{antila22b}. 
%
%For example, OPLS3e parameters overcome CHARMM36 in structural quality for POPC but predicts overestimated ion binding~\cite{kurki22}. On the other hand, CHARMM36 gives the best description for lipid headgroup conformational ensembles~\cite{bacle21}, while GROMOS-CKP parameters best capture the membrane packing in POPS bilayers~\cite{antila22b}. 
%
%Understanding of how closely simulations resemble the reality is crucial for selecting best models for specific applications,
The quality of lipid bilayer MD simulations has to be carefully assessed in their applications to minize the detrimental consequences of artificial MD simulation results for lipid bilayers~\cite{antila22b}. This is possible, for example, by using C--H bond order parameters from NMR and X-ray scattering form factors \cite{ollila16}, although it requires comparisons between large number of simulations which is laborious even with collaborative approaches~\cite{botan15,catte16,antila19,bacle21}. The quantitative quality measures in the NMRlipids databank substantially streamline this process by enabling automatic ranking of lipid bilayer simulations based on their quality against experiments. Here we demonstrate this by selecting the best models for simulations of membranes with POPC and POPE lipids that are the two most biologically abundant neutral membrane lipids~\cite{vanmeer08}.

Predictions for the lateral packing of POPC and POPE membranes in terms of area per lipid, one of the most important parameter used to characterize cellular membranes, diverge between different force fields as illustrated in Fig.~\ref{fig:quality}A. Because the membrane packing  correlates with the average order parameter of the $\textit{sn}$-1 chain (Fig.~\ref{fig:quality} F), simulations with the best quality for this part of POPE and POPC lipids are considered to give the most realistic results in Fig.~\ref{fig:POPCPOPEapls}C. To this end, the ranking based on C--H bond order parameter quality for the best performing POPC and POPE simulations in the NMRlipids databank are shown in Figs.~\ref{fig:quality}A and B, respectively. Direct comparisons with the experimental data are shown in Figs.~\ref{fig:quality}C--E and~\ref{fig:POPC_POPE_dataSI} for the most relevant simulations. Based on these rankings, simulations with Slipids parameters are the most realistic for POPC lipids with a clear difference to others, and 2\textsuperscript{nd} most realistic for POPE with only marginally lower quality than GROMOS-CKP simulations. Simulations with CHARMM36 and GROMOS-CKP parameters predict overly packed bilayer for POPC (overestimated order in Fig.~\ref{fig:POPC_POPE_dataSI}), while the area per lipid for POPC is overestimated in Lipid17. For POPE, CHARMM36 and Lipid17 predict membranes that are too packed. In conclusion, the quality evalution based on the NMRlipids databank suggests that Slipids parameters are the best available choice for simulations with PC and PE lipids, at least for the cases were membrane packing is relevant.

The C--H bond order parameters are sensitive to conformational ensembles of individual lipid molecules~\cite{ollila16}, while correlating well also with the membrane thickness and lateral packing as shown in Fig.~\ref{fig:quality}F. The X-ray scattering form factors are related to membrane dimensions \textit{via} the electron density profile. In particular, locations of minima in these form factors are correlated with the membrane lateral packing and thickness as shown in Figs.~\ref{fig:quality}F,~\ref{fig:QualityCorrelationsSI}, and~\ref{fig:sizedependence}. For quantitative quality measures, probabilities for each order parameter to locate within experimental error, and the distance of the first form factor minimum from the experimental value are calculated from a simulation.
%For detailed definitions of quality measures, see the supplementary information.
Together with the API, these enable rapid and automatic quality evaluation of simulations in the NMRlipids databank, thereby promoting the selection of best simulations for specific applications without laborious force field evaluation, and automatic parametrization procedures for atomistic and coarse grained simulations.

@@

The focus of biomolecular simulations is moving from studies of individual molecules to larger complexes and even whole cells and organelles~\cite{johnson15,thornburg22,gupta22}. Simultaneously, machine learning based models predicting behaviour of biomolecules and automatic approaches to parametrize models are emerging~\cite{jumper21,antila22b}. The resources delivered by the NMRlipids databank will support the development in both of these directions. The automatic quality evaluation and ranking of simulations against experimental data can be used to select the best parameters for simulations of large biomolecular complexes and to guide optimizations of new force field parameters for both atomistic and coarse grained simulations. The accuracy of simulations is becoming increasingly important due the accumulation of small errors when complexity and size of simulated systems are increasing.

@@

As a measure of equilibration quality for each trajectory, the ratio between the convergence time of first PC and the trajectory length was calculated for each lipid type individually. Sterols were excluded from the analysis. The script that calculates the equilibration of lipids is available at \texttt{Scripts/BuildDatabank/NMRPCA\_timerelax.py} in the NMRlipids databank repository. The resulting values are stored in files named \texttt{eq\_times.json} at folders in \texttt{/Data/Simulations} in the NMRlipids databank repository.

>>>>

Cellular membranes contain hundreds of different types of lipid molecules that regulate the membrane properties, morphology, and biological functions~\cite{vanmeer08,Lorent:2020a,Slatter:2016a}. Membrane lipid composition is implicated in diseases, such as cancer and neurodegenerative disorders, and therapeutics that affect membrane compositions are emerging~\cite{torres21}. However, biomembranes are often difficult to study experimentally because they are complex mitures of lipids and proteins that are in disordered fluid state with complex phase behaviour at biological conditions. For those reasons, the detailed connections between complex lipid interactions and biological functions taking place in or around membranes remain poorly understood. Molecular dynamics (MD) simulations have been particularly useful in understanding membrane systems, although their accuracy has often been compromised by artefacts such as the quality of model parameters~\cite{antila22b,gupta22}. Presently, the accuracy of models is becoming increasingly important as researches are progressing from simulations of individual molecules to simulating whole organelles or even cells using interdisciplinary approaches~\cite{johnson15,thornburg22,gupta22}. Such systems exhibit complex emergent behaviour making inaccuracies more difficult to detect and accumulation of even modest errors may have major impact on the conclusions.

@@

\subsection{NMRlipids databank-API: programmatic access to the MD simulation data}
The NMRlipids databank-API, available at \url{www.github.com/NMRLipids/Databank}, provides programmatic access to all simulation data in the NMRlipids databank. This enables wide range of novel data-driven applications from construction of machine learning models that predict membrane properties to automatic analyses of virtually any property from all simulations in the databank. Practical implementation of such analyses is illustrated in the flowchart in Fig.~\ref{fig:quality}D. First, raw data of each simulation can be accessed by iterating through the README.yaml files in the {\it databank layer} that contain links pointing to the locations of simulation trajectories and other relevant files. Simulations can be then automatically analyzed with the help of README.yaml and mapping files that affiliate specific naming conventions in each simulation to the universal molecule and atom names. Finally, the analysis results are stored using the same structure as in the {\it databank layer}.

@@

\subsection{Applications of the NMRlipids databank}
\subsubsection{Selection of simulation parameters using the NMRlipids databank: Best models for most abundant neutral membrane lipids}
%Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19,??}, but the lack of universal quality measure for membrane simulations complicates the estimations of reliability of simulations. 
%
%For example, OPLS3e parameters overcome CHARMM36 in structural quality for POPC but predicts overestimated ion binding~\cite{kurki22}. On the other hand, CHARMM36 gives the best description for lipid headgroup conformational ensembles~\cite{bacle21}, while GROMOS-CKP parameters best capture the membrane packing in POPS bilayers~\cite{antila22b}. 
%
The quality of lipid bilayer MD simulations has to be carefully assessed in their applications to minize the detrimental consequences of artificial MD simulation results for lipid bilayers~\cite{antila22b}. This is possible, for example, by using C--H bond order parameters from NMR and X-ray scattering form factors \cite{ollila16}, although it requires comparisons between large number of simulations which is laborious even with collaborative approaches~\cite{botan15,catte16,antila19,bacle21}. To streamline this process we have defined quantitative quality measures for lipid bilayer simulations that enable automatic ranking of lipid bilayer simulations based on their quality against experiments. Conformational ensembles of individual lipid molecules are evaluated in the NMRlipids databank by calculating the probabilities for each C-H bond order parameter to locate within experimental error. Furthermore, the quality against X-ray scattering experiments is estimated from the distance of the first form factor minimum between the experiment and simulation. These measures can be used to evaluate membrane dimensions as the form factor minima locations and acyl chain order parameters correlate with the membrane lateral packing and thickness as shown in Figs.~\ref{fig:quality}G,~\ref{fig:QualityCorrelationsSI}, and~\ref{fig:sizedependence}. 
%while correlating well also with the membrane thickness and lateral packing as shown in Fig.~\ref{fig:quality}F. 
%The X-ray scattering form factors are related to membrane dimensions \textit{via} the electron density profile. 
Here we demonstrate how automatic simulation quality evaluation and NMRlipids databank-API enable rapid selection of the best models for simulations of membranes with POPC and POPE lipids that are the two most biologically abundant neutral membrane lipids~\cite{vanmeer08}.

Predictions for the lateral packing of POPC and POPE membranes in terms of area per lipid, one of the most important parameter used to characterize cellular membranes, diverge between different force fields as illustrated in Fig.~\ref{fig:quality}A. To select the most realistic models among these, we ranked the simulations according to the quality of C-H bond order parameters evaluated against experimental data in Fig.~\ref{fig:top50simulations}. Because the membrane packing correlates with the average order parameter of the $\textit{sn}$-1 chain (Fig.~\ref{fig:quality}G), force fields occurring in Fig.~\ref{fig:quality}A are ranked according to the $\textit{sn}$-1 chain quality in Figs.~\ref{fig:quality}B and C. Also the highest ranked simulation (POPC bilayer with OPLS3e parameters) is included for the reference.
%for this part of POPE and POPC lipids are considered to give the most realistic results in Fig.~\ref{fig:POPCPOPEapls}C. To this end, the ranking based on C--H bond order parameter quality for the best performing POPC and POPE simulations in the NMRlipids databank are shown in Figs.~\ref{fig:quality}A and B, respectively. 
Direct comparisons with the experimental data are shown in Figs.~\ref{fig:quality}C--E and~\ref{fig:POPC_POPE_dataSI} for the most relevant simulations. Based on these rankings, Slipids simulation gives the most realistic prediction for POPC membrane in Fig.~\ref{fig:quality}A with a clear difference to others, and 2\textsuperscript{nd} most realistic prediction for POPE with only marginally lower quality than GROMOS-CKP simulations. Simulations with CHARMM36 and GROMOS-CKP parameters predict overly packed bilayer for POPC (overestimated order in Fig.~\ref{fig:POPC_POPE_dataSI}), while the area per lipid for POPC is overestimated in Lipid17. For POPE, CHARMM36 and Lipid17 predict membranes that are too packed. In conclusion, the quality evaluation based on the NMRlipids databank suggests that Slipids parameters are the best available choice for simulations with PC and PE lipids, at least for applications where membrane packing is relevant.

@@

The focus of biomolecular simulations is moving from studies of individual molecules to larger complexes and even whole cells and organelles~\cite{johnson15,thornburg22,gupta22}. Simultaneously, machine learning based models predicting behaviour of biomolecules and automatic approaches to parametrize models are emerging~\cite{jumper21,antila22b}. The resources delivered by the NMRlipids databank will support the development in both of these directions. Automatic quality evaluation and ranking of simulations against experimental data enables the selection of best simulations for specific applications without laborious manual force field evaluation. This also streamlines automatic parametrization procedures for atomistic and coarse grained simulations.
%can be used to select the best parameters for simulations of large biomolecular complexes and to guide optimizations of new force field parameters for both atomistic and coarse grained simulations. 
Such practises fostering the accuracy of simulations are becoming increasingly important due the accumulation of small errors when complexity and size of simulated systems are increasing.

@@

As a measure of equilibration quality for each trajectory, the ratio between the convergence time of first PC and the trajectory length, $\tau_{\rm rel} = \tau_{\text{convergence}}/\tau_{\text{sim}}$, was calculated for each lipid type individually. Sterols were excluded from the analysis. The script that calculates the equilibration of lipids is available at \texttt{Scripts/BuildDatabank/NMRPCA\_timerelax.py} in the NMRlipids databank repository. The resulting values are stored in files named \texttt{eq\_times.json} at folders in \texttt{/Data/Simulations} in the NMRlipids databank repository.
